Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-20 paper/paper.tex
Typos

\section{Introduction}
\label{sec:intro}
Structured recursion schemes are powerful abstractions that capture common
patterns of recursion. The main benefit of structuring computation using
recursion schemes, is that they enjoy well-estabished \emph{algebraic
properties} that can serve as a foundation for reasoning about program
equivalences, transformations, and optimisations (e.g.\ \emph{fusion} laws, or
semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
%
These algebraic properties are widely used in the context of \emph{program
calculation}, where programmers describe their code using simple, inefficient
specifications in an \emph{algebra of programming}~\cite{BirddeMoor96:Algebra},
and then use the algebraic laws of this algebra of programming to
\emph{calculate} an efficient version of the same algorithm.
%
Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml
implementation, with
\ocaml{map (fun x -> x * 2)}.
A lot of program calculations have this pattern: start from a simple
specification, e.g.\ $\mathsf{map}\;(\lambda x.\;x \times 2) \circ
\mathsf{sort}$, and use program equivalences and algebraic laws to rewrite it to
an optimised version. We will revisit a similar example in
Section~\ref{sec:sorting}.

This program calculation stems from the more general theory of hylomorphisms.  A
hylomorphism is a structured recursion scheme representing the idea of a
divide-and-conquer algorithm where, first, the problem is split into smaller
subproblems (divide), then subsolutions are computed recursively, and, finally,
these subsolutions are put together to compute the solution to the original
problem (conquer).  This general pattern of recursion induces a reasoning
principle, called \emph{fusion law}, which has been exemplified in the previous
paragraph.

@@

An example of a recursive coalgebra is the \coq{partition} function in quicksort
which destructures a list into a pivot and two sublist and as long as the
sublists are smaller the partitioning function still yields a unique solution to
the recursion scheme.  The uniqueness property of the hylo yields the following
fusion laws:
\begin{align}
  \label{eq:hylo:fusion}
  f \circ \hylo{c}{a} & = \hylo{c}{a'}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  f \circ a = a' \circ F(f) \\
  \hylo{c}{a} \circ f & = \hylo{c'}{a}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  c \circ f = F(f) \circ c'
\end{align}
Using the hylo fusion laws, we can prove the well-known \emph{deforestation
optimisation}, also known as the \emph{composition
law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
computations, one that builds a data structure, and another one that consumes
it, can be fused together into a single recursive definition. This, in turn,
allows us to prove that a recursive hylomorphism is the composition of a
catamorphism and a recursive anamorphism.

@@

Recall that an algebra is a set $A$ together with a morphism that defines the
operations of the algebra $F\;A \to A$.  Given a type \coq{A} and a container
\coq{F}, an `\coq{App F}'-algebra (or `\coq{F}-algebra' for short) is a pair
given by the carrier \coq{A}, and the \emph{structure map} of type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The initial \coq{F}-algebra can be defined as the inductive type with a finite
number of \coq{App F}
\begin{coqcode}
Inductive LFix F : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
\coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its
inverse $\InOp$ (see Section~\ref{sec:category-theory}).  As an example, the
initial \coq{F}-algebra for the container that is isomorpic to the functor
\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra being
defined by the empty list \coq{Empty : unit -> LFix F} and the cons operation
\coq{Cons : A * LFix F -> LFix F}.

@@

In order to find a solution we restrict ourselves to the so-called
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  We mechanise
\emph{recursive hylomorphisms} which are guaranteed to have a unique solution to
the hylomorphism equation. These are hylomoprhisms where the coalgebra is
\emph{recursive}, i.e.\ coalgebras that terminate on all inputs. We represent
recursive coalgebras using a predicate that states that a coalgebra terminates
on an input:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
For convenience, we package recursive coalgebras with proofs that they terminate
for all inputs:
\begin{coqcode}
Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
\end{coqcode}
Recursive hylomorphisms are implemented in Coq recursively on the structure of
the proof that the anamorphism terminates on an input:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) : forall (x : A), RecF c x -> B
  := fix f x H
     := match c x as cx return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B with
        | cx => fun H => a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
        end (RecF_inv H).
\end{coqcode}

@@

\section{Extraction}
\label{sec:examples}
We go in this section through a series of case studies of various recursive
algorithms.  We show how they can be encoded in our framework, how can we do
program calculation techniques for optimisation, and how can they be extracted
to idiomatic functional code. Although our examples use OCaml as target
language, extraction works equally well to other target languages. Our examples
are the Quicksort and Mergesort algorithms (Section~\ref{sec:sorting}), dynammic
programming and Knapsack (Section~\ref{sec:knap}), and examples of the shortcut
deforestation optimisation in our framework (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

@@

\subsection{Shortcut Deforestation}\label{sec:shortcut}
The final case study we consider is shortcut deforestation on lists.  Shortcut
deforestation can be expressed succintly in terms of hylomorphisms and their
laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
However, we prove a specific version of this theorem for the container whose
least fixed point is a list, and use it to encode the example from Takano and
Meijer~\cite{TakanoM95}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here, we are defining function \coq{sf1} as the composition of \coq{length},
\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
catamorphisms. Function \coq{append ys} is also a catamorphism that
appends \coq{ys} to an input list. It is defined by applying an algebra to
every cons node of a list, and applying a catamorphism with the input algebra
to \coq{ys} in the nil case:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
\end{coqcode}
Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
definitions, and simplify in our specification for \coq{sf1}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
  rewrite hylo_map_fusion, <- acid_rain. simpl; reflexivity.
Defined.
\end{coqcode}
From this, we extract the following OCaml code:
\begin{ocamlcode}
let rec sf1 f ys =
  function | [] -> let rec f0 = function
                   | [] -> (Uint63.of_int (0))
                   | _ :: t -> add (Uint63.of_int (1)) (f0 t)
                   in f0 ys
           | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
\end{ocamlcode}

@@

% \textbf{Program Calculation}.
% \label{sec:calccoq}
Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs from
their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. Programmers may specify algorithms
abstractly at a high level, and then derive an efficient implementation using
the process of program calculation~\cite{BirddeMoor96:Algebra}. This area has
been extensively explored over the years.

@@

%\textbf{Termination Checking of Nonstructural Recursion}.% \label{sec:termcheck}
% Termination and productivity are non-trivial properties of programs, and various
% methods have been proposed for checking that these properties hold. A common
% approach is guardedness checking~\cite{Eduardo94}, which is a
% \emph{syntactic} check that definitions avoid the introduction of
% non-normalizable terms. This sort of check generally looks for a
% \emph{structural decrease} of arguments in the recursive calls of a function.
% Coq uses such a check, and it works for many classic functional programming
% patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
% definitions will not be accepted by such checks.
%In particular,
The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.

>>>>

\section{Introduction}
\label{sec:intro}
Structured recursion schemes are powerful abstractions that capture common
patterns of recursion. The main benefit of structuring computation using
recursion schemes, is that they enjoy well-established \emph{algebraic
properties} that can serve as a foundation for reasoning about program
equivalences, transformations, and optimisations (e.g.\ \emph{fusion} laws, or
semi-automatic
parallelisations~\cite{TakanoM95,Gibbons96:Third,Morihata09:Third,farmsCastro}).
%
These algebraic properties are widely used in the context of \emph{program
calculation}, where programmers describe their code using simple, inefficient
specifications in an \emph{algebra of programming}~\cite{BirddeMoor96:Algebra},
and then use the algebraic laws of this algebra of programming to
\emph{calculate} an efficient version of the same algorithm.
%
Suppose, for example, that we want to write a program that sorts a list of
integers, and then multiplies by 2 all its elements. In OCaml, we may write this
function directly:
\begin{minted}{ocaml}
let rec sort_times_two = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x -> x < h) t in
  sort_times_two l @ (h * 2) :: sort_times_two t
\end{minted}
This program is derived by \emph{fusing} a regular \emph{quicksort} OCaml
implementation, with
\ocaml{map (fun x -> x * 2)}.
A lot of program calculations have this pattern: start from a simple
specification, e.g.\ $\mathsf{map}\;(\lambda x.\;x \times 2) \circ
\mathsf{sort}$, and use program equivalences and algebraic laws to rewrite it to
an optimised version. We will revisit a similar example in
Section~\ref{sec:sorting}.

This program calculation stems from the more general theory of hylomorphisms.  A
hylomorphism is a structured recursion scheme representing the idea of a
divide-and-conquer algorithm where, first, the problem is split into smaller
sub-problems (divide), then sub-solutions are computed recursively, and, finally,
these sub-solutions are put together to compute the solution to the original
problem (conquer).  This general pattern of recursion induces a reasoning
principle, called \emph{fusion law}, which has been exemplified in the previous
paragraph.

@@

An example of a recursive coalgebra is the \coq{partition} function in quicksort
which destructures a list into a pivot and two sub-list and as long as the
sub-lists are smaller the partitioning function still yields a unique solution to
the recursion scheme.  The uniqueness property of the hylomorphisms yields the following
fusion laws:
\begin{align}
  \label{eq:hylo:fusion}
  f \circ \hylo{c}{a} & = \hylo{c}{a'}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  f \circ a = a' \circ F(f) \\
  \hylo{c}{a} \circ f & = \hylo{c'}{a}
  \hspace{1cm}
  \Longleftarrow
  \hspace{1cm}
  c \circ f = F(f) \circ c'
\end{align}
Using the hylomorphism fusion laws, we can prove the well-known \emph{deforestation
optimisation}, also known as the \emph{composition
law}~\cite{DBLP:conf/ifl/HinzeHJ10}. This is when two consecutive recursive
computations, one that builds a data structure, and another one that consumes
it, can be fused together into a single recursive definition. This, in turn,
allows us to prove that a recursive hylomorphism is the composition of a
catamorphism and a recursive anamorphism.

@@

Recall that an algebra is a set $A$ together with a morphism that defines the
operations of the algebra $F\;A \to A$.  Given a type \coq{A} and a container
\coq{F}, an `\coq{App F}'-algebra (or `\coq{F}-algebra' for short) is a pair
given by the carrier \coq{A}, and the \emph{structure map} of type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The initial \coq{F}-algebra can be defined as the inductive type with a finite
number of \coq{App F}
\begin{coqcode}
Inductive LFix F : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
\coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its
inverse $\InOp$ (see Section~\ref{sec:category-theory}).  As an example, the
initial \coq{F}-algebra for the container that is isomorphic to the functor
\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra being
defined by the empty list \coq{Empty : unit -> LFix F} and the cons operation
\coq{Cons : A * LFix F -> LFix F}.

@@

In order to find a solution we restrict ourselves to the so-called
\emph{recursive coalgebras}~\cite{AdamekMM19,CaprettaUV04}.  We mechanise
\emph{recursive hylomorphisms} which are guaranteed to have a unique solution to
the hylomorphism equation. These are hylomorphisms where the coalgebra is
\emph{recursive}, i.e.\ coalgebras that terminate on all inputs. We represent
recursive coalgebras using a predicate that states that a coalgebra terminates
on an input:
\begin{coqcode}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{coqcode}
For convenience, we package recursive coalgebras with proofs that they terminate
for all inputs:
\begin{coqcode}
Notation RCoalg F A := ({ c : Coalg F A | forall x, RecF c x }).
\end{coqcode}
Recursive hylomorphisms are implemented in Coq recursively on the structure of
the proof that the anamorphism terminates on an input:
\begin{coqcode}
Definition hylo_def (a : Alg F B) (c : Coalg F A) : forall (x : A), RecF c x -> B
  := fix f x H
     := match c x as cx return (forall e : Pos (shape cx), RecF c (cont cx e)) -> B with
        | cx => fun H => a { shape := shape cx ; cont := fun e => f (cont cx e) (H e) }
        end (RecF_inv H).
\end{coqcode}

@@

\section{Extraction}
\label{sec:examples}
We go in this section through a series of case studies of various recursive
algorithms.  We show how they can be encoded in our framework, how can we do
program calculation techniques for optimisation, and how can they be extracted
to idiomatic functional code. Although our examples use OCaml as target
language, extraction works equally well to other target languages. Our examples
are the Quicksort and Mergesort algorithms (Section~\ref{sec:sorting}), dynamic
programming and Knapsack (Section~\ref{sec:knap}), and examples of the shortcut
deforestation optimisation in our framework (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

@@

\subsection{Shortcut Deforestation}\label{sec:shortcut}
The final case study we consider is shortcut deforestation on lists.  Shortcut
deforestation can be expressed succinctly in terms of hylomorphisms and their
laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
However, we prove a specific version of this theorem for the container whose
least fixed point is a list, and use it to encode the example from Takano and
Meijer~\cite{TakanoM95}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here, we are defining function \coq{sf1} as the composition of \coq{length},
\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
catamorphisms. Function \coq{append ys} is also a catamorphism that
appends \coq{ys} to an input list. It is defined by applying an algebra to
every cons node of a list, and applying a catamorphism with the input algebra
to \coq{ys} in the nil case:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
\end{coqcode}
Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
definitions, and simplify in our specification for \coq{sf1}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
  rewrite hylo_map_fusion, <- acid_rain. simpl; reflexivity.
Defined.
\end{coqcode}
From this, we extract the following OCaml code:
\begin{ocamlcode}
let rec sf1 f ys =
  function | [] -> let rec f0 = function
                   | [] -> (Uint63.of_int (0))
                   | _ :: t -> add (Uint63.of_int (1)) (f0 t)
                   in f0 ys
           | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
\end{ocamlcode}

@@

% \textbf{Program Calculation}.
% \label{sec:calccoq}
Within the domain of program optimisation, program calculation serves as a
well-established programming technique aimed at deriving efficient programs from
their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. Programmers may specify algorithms
abstractly at a high level, and then derive an efficient implementation using
the process of program calculation~\cite{BirddeMoor96:Algebra}. This area has
been extensively explored over the years.

@@

%\textbf{Termination Checking of Nonstructural Recursion}.% \label{sec:termcheck}
% Termination and productivity are non-trivial properties of programs, and various
% methods have been proposed for checking that these properties hold. A common
% approach is guardedness checking~\cite{Eduardo94}, which is a
% \emph{syntactic} check that definitions avoid the introduction of
% non-normalizable terms. This sort of check generally looks for a
% \emph{structural decrease} of arguments in the recursive calls of a function.
% Coq uses such a check, and it works for many classic functional programming
% patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
% definitions will not be accepted by such checks.
%In particular,
The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of division found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognisable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.
